Modeling the effect of aneuploidy on cancer evolution

### 2023-02-19 EvolutionaryRescue1.tex
remus update feb 17

\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{tikz,pgfplots}
\usepackage{gensymb}
\usepackage{graphicx}

\usepackage{epstopdf}
%\usepackage{listings}
\usepackage{tikz,fp,fp-random,todonotes}
\usepackage[procnames]{listings}
\usepackage{afterpage}
\usepackage{float}
\usepackage{amsthm}
\usepackage{wrapfig}
\usepackage{booktabs,caption}
\usepackage[flushleft]{threeparttable}
\usepackage{scrextend}
\usepackage[all]{nowidow}
\usepackage{stfloats}
\usepackage{mathtools}
\usepackage{bm}
\usepackage{stackengine}
\usepackage[hidelinks]{hyperref}
\usepackage{siunitx}
\usepackage{subfig}
\usepackage{authblk}
\usepackage{cleveref}

@@

% line numbers
\usepackage[displaymath, mathlines]{lineno}
\renewcommand\linenumberfont{\normalfont\small\sffamily}
\linenumbers
\modulolinenumbers[2]

@@

% TODO remove the commented out paragraphs if you dont need them.

%The question of what role does chromosomal instability (CIN) play in the emergence of cancer has been studied extensively in the past decades \cite{michor2005can,christine2018understanding,nowak2002role,pavelka2010dr,komarova2003mutation,zhu2018cellular}. One hypothesis is that CIN facilitates tumorigenesis by accelerating the removal of tumor suppression genes (TSG) and subsequent appearance of cancer. The deletion of tumor suppression genes can happen in two ways (see Figure \ref{figureCIN}): two point mutations deleting both alleles of the TSG or one point mutation and one chromosomal loss event. The case of two chromosomal loss events is taken into consideration because the fitness cost would be very large and the cell would die.
%
%Populations adapted to a certain environment are vulnerable to change in the environment which might cause extinction of the population. Adaptation is a race against time as the population size decreases in the new environment. Evolutionary rescue is the process where the population acquires mutations which increases the fitness of the population in the new environment such that the extinction in averted. We are interested in how likely is a population to adapt to the changing environment through evolution. Examples of such changes include climate change, invasive species or the onset of drug therapies. There exists three possible ways for a population to survive environmental degradation: migration to a new habitat similar to the one before the onset of environmental changes \cite{cobbold2020should}, adaptation by phenotypic plasticity which involves no changes in the genotype\cite{carja2019evolutionary,carja2017evolutionary} and adaptation through genetic mutations\cite{uecker2014evolutionary,uecker2016role,uecker2011fixation}.
%
%Majority of studies assume that the fitness of the wildtype and mutant are time homogeneous. An exception is \cite{marrec2020adapt} fitness of wildtype and mutant are taken to be functions of time.

@@

\paragraph{Aneuploidy in cancer.} Chromosomal instability (CIN) is the mitotic process in which cells suffer from chromosome mis-segregation that leads to aneuploidy, where cells are characterized by structural changes of the chromosomes and copy number alterations \cite{schukken2018cin}.
Interestingly, aberrations in chromosome copy number are have been shown to allow cancer cells to survive under stressful conditions such as drug therapy. % REF
Indeed, cancer cells are often likely to be aneuploid, and aneuploidy is associated with poor patient outcomes \cite{ben2020context}.

@@

The role of chromosomal instability (CIN) in the emergence of cancer has been studied extensively in the past decades \cite{michor2005can,christine2018understanding,nowak2002role,pavelka2010dr,komarova2003mutation,zhu2018cellular}.
One hypothesis is that CIN facilitates tumor genesis by accelerating the removal of tumor suppression genes (TSG) and subsequent appearance of cancer. The deletion of tumor suppression genes can happen in two ways: two point mutations deleting both alleles of the TSG (assuming a diploid genotype), or one point mutation and one chromosomal loss event.
Initial studies have shown that aneuploidy can have a significant role in the deletion of the the tumor suppressing genes when compared to two consecutive point mutations\cite{nowak2002role,komarova2003mutation,michor2005can,komarova2008selective}. % TODO mention if studies are theoretical , experimental, or clinical
However, when taking into account that the appearance of aneuploidy requires a mutation to trigger CIN, the probability that CIN precedes tumor genesis is highly unlikely.

@@

The difference between growth and death is $\Delta_k = \lambda_k-\mu_k$.
We assume that $\Delta_w<0$ due to drug therapy (or some other stress), and that $\Delta_m$ due to resistance conferred by the mutant.
We analyze two cases: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, % REF: https://www.nature.com/articles/nrmicro.2016.34
$0>\Delta_a>\Delta_w$.

% TODO explain briefly what evolutionary rescue means in this model.
To analyze evolutionary rescue in this model, we use the framework of multitype branching processes. % REF Harris? https://doi.org/10.1016/j.jtbi.2021.110849
This allows us to find explicit expressions for the survival probability of the population--the probability that it does not become extinct due to stressful conditions, e.g., drug therapy.
Note that branching processes imply that lineages produced by cells from the initial population grow and die independently of each other. This is a necessary simplification that is not biologically realistic, as cells usually compete for resources. A more biologically realistic model... % add the description of the more realistic model here

@@

% TODO add subsection on the Gillespie simulations

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TODO combine all diagram figures to a single figure with multiple panels. and make it fit in a single page.
\begin{figure}
\centering
\begin{tikzpicture}[very thick,decoration={
    markings,
    mark=at position 0.5 with {\arrow{>}}}
    ] 
\edef\x{6.5} 
\edef\y{-0.5} 
\edef\r{1.25}

\draw [postaction={decorate}, thick](-0.25,-2) -- (-0.25+5,0) node [midway, above] {$2u$};
\draw [postaction={decorate}, thick](-0.25,-2) -- (-0.25+5,-4) node [midway, above] {$2p$};
\draw [postaction={decorate}, thick](-0.25+5+\r,-0) -- (-0.25+10-\r,-0) node [midway, above] {$u$};
\draw [postaction={decorate}, thick](-0.25+5+\r,-4) -- (-0.25+10-\r,-4) node [midway, above] {$u$};
\draw [postaction={decorate}, thick](-0.25+5,-0) -- (-0.25+10,-4) node [midway, above] {$p$};

\filldraw[color=red!60, fill=red!5, very thick](-0.25,-2) circle (\r);
\filldraw[color=red!60, fill=red!5, very thick](-0.25+5,0) circle (\r);
\filldraw[color=red!60, fill=red!5, very thick](-0.25+5,-4) circle (\r);
\filldraw[color=red!60, fill=red!5, very thick](-0.25+10,0) circle (\r);
\filldraw[color=red!60, fill=red!5, very thick](-0.25+10,-4) circle (\r);

\draw [thick](0.05,-1.25) -- (0.05,-2.75);
\draw [thick](-0.55,-1.25) -- (-0.55,-2.75);

\draw [thick](0.05+5,-1.25+2) -- (0.05+5,-2.75+2);
\draw [very thick, blue](-0.55+5,-1.25+2) -- (-0.55+5,-2.75+2);

\draw [very thick, blue](0.05+10,-1.25+2) -- (0.05+10,-2.75+2);
\draw [very thick, blue](-0.55+10,-1.25+2) -- (-0.55+10,-2.75+2);

\draw [thick](-0.55+5,-1.25-2) -- (-0.55+5,-2.75-2);

\draw [very thick, blue](-0.55+10,-1.25-2) -- (-0.55+10,-2.75-2);
\end{tikzpicture}
\caption{Diagram of tumorigenesis by which a cell loses both of its tumor suppressing cells. The first tumor suppressing gene can be eliminated by a mutation at rate $2u$ or through chromosomal instability with rate $2p$. The last TSG can either be eliminated through mutation at rate $u$ or through chromosomal instability with rate $p$. Grey vertical lines represent active TSGs while the blue vertical lines represent inactivated TSGs.
}
\label{figureCIN}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}
\centering
\begin{tikzpicture}[very thick,decoration={
    markings,
    mark=at position 0.5 with {\arrow{>}}}
    ] 
\edef\x{6.5} 
\edef\y{-0.5} 
\edef\r{1.25}

\draw [postaction={decorate}, thick](-0.25,-4) -- (-0.25+5,0) node [midway, above] {$u$};
%\draw [postaction={decorate}, thick](-0.25,-2) -- (-0.25+5,-4) node [midway, above] {$2p$};
\draw [postaction={decorate}, thick](-0.25+5,-0) -- (-0.25+10,-4) node [midway, above] {$v$};
%\draw [postaction={decorate}, thick](-0.25+5+\r,-4) -- (-0.25+10-\r,-4) node [midway, above] {$u$};
%\draw [postaction={decorate}, thick](-0.25+5,-0) -- (-0.25+10,-4) node [midway, above] {$p$};
\draw [postaction={decorate}, thick]  (-0.25+10,-4) -- (-0.25+5,-8) node [midway, above] {$u$};
\draw [postaction={decorate}, thick](-0.25,-4)  -- (-0.25+5,-8)  node [midway, above] {$v$};

\filldraw[color=red!60, fill=red!5, very thick](-0.25,-4) circle (\r);
\filldraw[color=red!60, fill=red!5, very thick](-0.25+5,0) circle (\r);
%\filldraw[color=red!60, fill=red!5, very thick](-0.25+5,-4) circle (\r);
\filldraw[color=red!60, fill=red!5, very thick](-0.25+10,-4) circle (\r);
%\filldraw[color=red!60, fill=red!5, very thick](-0.25+10,-4) circle (\r);
\filldraw[color=red!60, fill=red!5, very thick](-0.25+5,-8) circle (\r);

\draw [thick](0.05,-1.25+2-4) -- (0.05,-2.75+2-4);
\draw [thick](-0.55,-1.25+2-4) -- (-0.55,-2.75+2-4);

\draw [thick](0.15+5,-1.25+2) -- (0.15+5,-2.75+2);
\draw [thick](-0.65+5,-1.25+2) -- (-0.65+5,-2.75+2);
\draw [thick](-0.25+5,-1.25+2) -- (-0.25+5,-2.75+2);

\draw [thick](0.15+10,-1.25+2-4) -- (0.15+10,-2.75+2-4);
\draw [thick](-0.25+10,-1.25+2-4) -- (-0.25+10,-2.75+2-4);
\draw [color=blue,very thick](-0.7+10,-1.25+2-4) -- (-0.7+10,-2.75+2-4);

\draw [thick](0.05+5,-1.25+2-8) -- (0.05+5,-2.75+2-8);
\draw [color=blue,very thick](-0.55+5,-1.25+2-8) -- (-0.55+5,-2.75+2-8);
%\draw [thick](-0.55+5,-1.25-2) -- (-0.55+5,-2.75-2);

%\draw [thick](-0.55+10,-1.25-2) -- (-0.55+10,-2.75-2);
\end{tikzpicture}
\caption{Diagram of the evolutionary rescue process which involves aneuploidy. The wildtype cell can either acquire trisomy at rate $u$ or acquire a fitness increasing mutation at rate $v$. Aneuploidy cells can  acquire a fitness increasing mutation at rate $v$. Finally, the aneuploidy cells can lose trisomy at rate $u$.
}
\label{figuretrisomy}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\centering
\begin{tikzpicture}[very thick,decoration={
    markings,
    mark=at position 0.5 with {\arrow{>}}}, middlearrow/.style 2 args={
        decoration={             
            markings, 
            mark=at position 0.5 with {\arrow[xshift=3.333pt]{triangle 45}, \node[#1] {#2};}
        },
        postaction={decorate}
    }]
\edef\x{6.5} 
\edef\y{-0.5} 
\edef\r{1.25}

\edef\R{1} 
\edef\Y{0.25}

\draw[black,thick,middlearrow={above}{$\lambda_w$},text=black] (-0.25,\Y) arc (270:-90:\R);
\draw[black,thick,middlearrow={above}{$\lambda_a$},text=black] (-0.25+5,\Y) arc (270:-90:\R);
\draw[black,thick,middlearrow={above}{$\lambda_m$},text=black] (-0.25+10,\Y) arc (270:-90:\R);

\draw [postaction={decorate}, thick](-0.25,0) -- (-0.25+5,0) node [midway, above] {$u$};
\draw [postaction={decorate}, thick](-0.25+5+\r,-0) -- (-0.25+10-\r,-0) node [midway, above] {$v$};

\draw[thick, postaction={decorate}] (-0.25,0) arc (180:360:5) node [midway, above] {$v$};

\filldraw[color=green!60, fill=green!5, very thick](-0.25,0) circle (\r);
\filldraw[color=green!60, fill=green!5, very thick](-0.25+5,0) circle (\r);
\filldraw[color=green!60, fill=green!5, very thick](-0.25+10,0) circle (\r);

\draw [thick, postaction={decorate}](-0.25,-1.25) -- (-0.25,-2.75) node [midway, left] {$\mu_w$};

\draw [thick, postaction={decorate}](-0.25+5,-1.25) -- (-0.25+5,-2.75) node [midway, right] {$\mu_a$};

\draw [thick, postaction={decorate}](-0.25+10,-1.25) -- (-0.25+10,-2.75) node [midway, right] {$\mu_m$};

\edef\x{0.25} 
\edef\y{-0.25}

\node[text width=2cm,align=justify] at (\y,\x)  
{\begin{center}
Wildtype  euploid
\end{center}};

\node[text width=2cm,align=justify] at (\y+5,\x)  
{\begin{center}
Wildtype  aneuploid
\end{center}};

\node[text width=2cm,align=justify] at (\y+10,\x)  
{\begin{center}
Mutant  euploid
\end{center}};
\end{tikzpicture}
\caption{Diagram of the evolutionary rescue model where the population of cancer cells is subdivided to wildtype euploid, wildtype aneuploid and mutant euploid cells that divide and die with rates $\lambda_w,  \lambda_a,  \lambda_m$ and $\mu_w, \mu_a, \mu_m$, respectively. Wildtype cells can acquire aneuploidy at rate $u$. Aneuploid and wildtype cells can mutate with rate $v$.}
\label{figureAneuploidy}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\section*{Results}

@@

We write \cref{survprobw} as
\begin{equation}\label{survprobwapprox}
p_w=-\frac{1}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right).
\end{equation}
So, given an initial population of $N$ wildtype cells, the probability that the population will survive is given by
\begin{align}
p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right] .
\end{align}
\Cref{SurvPlot} shows that the survival probability, $p_{est}$, quickly grows as the wildtype growth rate, $\lambda_w$, increases. % Todo I dont think this figure should show lambda_w>1, as in that case the wildtype is resistant and there is no rescue scenario.
\Cref{SurvPlotNData} show $p_{est}$ as a function of $N$, including comparison of our approximation % ref to equation
and simulation results.

@@

\paragraph{$\lambda_a<\mu_a$ and $\lambda_w<\mu_w$.}
We assume that $\lambda_a<\mu_a$ and $\lambda_w<\mu_w$ and, as a result, we rewrite \eqref{survprobw} and \eqref{survproba} as:
\begin{align*}
p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1-\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right),\\
p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right).
\end{align*}
\begin{align}\label{survprobwinitial}
p_w&\approx-\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
\nonumber
&\approx\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}-\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ \label{survprobw2}
&\approx\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_w-\mu_w\right)}\left[\frac{u}{\left(\lambda_a-\mu_a\right)}-1\right],
\end{align}
where in the last line we have used the fact that $u,v\ll1$. Using the notational convention
\begin{equation}\label{notationalconv}
\Delta_i=\lambda_i-\mu_i,
\end{equation}
we write \eqref{survprobw} as
\begin{equation}\label{survprobwapprox}
p_w=\frac{v\Delta_m}{\lambda_m\Delta_w}\left(\frac{u}{\Delta_a}-1\right).
\end{equation}
Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
\begin{align}\label{AneuploidyDeleteriousApprox}
p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right],
\end{align}

@@

Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
\begin{align}\label{pestthirdcase}
p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right],
\end{align}
which we plot in Figure \ref{DeleteriousPlot} where we compare with numerical simulations and the exact result \eqref{survprobw}.

\begin{align}
\Delta_a^*=2vp_m+v+2\sqrt{vp_m\left(vp_m+\mu_a+v\right)}
\end{align}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!t]
 \vspace*{1\baselineskip}
\includegraphics[width=1\textwidth]{Figures/SurvPlotNData.pdf}
\caption{Plot of the probability of survival of a population as a function of the initial population size of wildtype cells. The blue line represents the exact solution \eqref{survprobw}, the orange line line represents the approximation \eqref{survprobwapprox}, the green line represents the first order correction \eqref{survprobwapproxcorrected} and the red dots represents stochastic simulations. For the simulations we have chosen the following parameters: $\lambda_w=1-10^{-2}, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1.$}
\label{SurvPlotNData}
\end{figure}

>>>>

\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{tikz,pgfplots}
\usepackage{gensymb}
\usepackage{graphicx}
\usepackage{subcaption}

\usepackage{epstopdf}
%\usepackage{listings}
\usepackage{tikz,fp,fp-random,todonotes}
\usepackage[procnames]{listings}
\usepackage{afterpage}
\usepackage{float}
\usepackage{amsthm}
\usepackage{wrapfig}
\usepackage{booktabs,caption}
\usepackage[flushleft]{threeparttable}
\usepackage{scrextend}
\usepackage[all]{nowidow}
\usepackage{stfloats}
\usepackage{mathtools}
\usepackage{bm}
\usepackage{stackengine}
\usepackage[hidelinks]{hyperref}
\usepackage{siunitx}
%\usepackage{subfig}
\usepackage{authblk}
\usepackage{cleveref}

@@

% line numbers
\usepackage[displaymath, mathlines]{lineno}
\renewcommand\linenumberfont{\normalfont\small\sffamily}
%\linenumbers
\modulolinenumbers[2]

@@

% TODO remove the commented out paragraphs if you dont need them. OK

@@

\paragraph{Aneuploidy in cancer.} Chromosomal instability (CIN) is the mitotic process in which cells suffer from chromosome mis-segregation that leads to aneuploidy, where cells are characterized by structural changes of the chromosomes and copy number alterations \cite{schukken2018cin}.
Interestingly, aberrations in chromosome copy number have been shown to allow cancer cells to survive under stressful conditions such as drug therapy. % REF
Indeed, cancer cells are often likely to be aneuploid, and aneuploidy is associated with poor patient outcomes \cite{ben2020context}.

@@

The role of chromosomal instability (CIN) in the emergence of cancer has been studied extensively in the past decades \cite{michor2005can,christine2018understanding,nowak2002role,pavelka2010dr,komarova2003mutation,zhu2018cellular}.
One hypothesis is that CIN facilitates tumor genesis by accelerating the removal of tumor suppression genes (TSG) and subsequent appearance of cancer. The deletion of tumor suppression genes can happen in two ways: two point mutations deleting both alleles of the TSG (assuming a diploid genotype), or one point mutation and one chromosomal loss event.
Initial theoretical studies have shown that aneuploidy can have a significant role in the deletion of the the tumor suppressing genes when compared to two consecutive point mutations\cite{nowak2002role,komarova2003mutation,michor2005can,komarova2008selective}. % TODO mention if studies are theoretical , experimental, or clinical OK
However, when taking into account that the appearance of aneuploidy requires a mutation to trigger CIN, the probability that CIN precedes tumor genesis is highly unlikely.

@@

The difference between growth and death is $\Delta_k = \lambda_k-\mu_k$.
We assume that $\Delta_w<0$ due to drug therapy (or some other stress), and that $\Delta_m$ due to resistance conferred by the mutant.
We analyze two cases: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, % REF: https://www.nature.com/articles/nrmicro.2016.34 OK
$0>\Delta_a>\Delta_w$  \cite{brauner2016distinguishing}.

% TODO explain briefly what evolutionary rescue means in this model. OK
In our model rescues means that the aneuploid cells are able to either reverse ($\Delta a>0$ case) or delay ($0>\Delta_a>\Delta_w$ case) the extinction of the cancer cell population until mutant cells are generated and rescue the population.
To analyze evolutionary rescue in this model, we use the framework of multitype branching processes \cite{rybnikov2021fitness,harris1963theory}. % REF Harris? https://doi.org/10.1016/j.jtbi.2021.110849 OK
This allows us to find explicit expressions for the survival probability of the population--the probability that it does not become extinct due to stressful conditions, e.g., drug therapy.
Note that branching processes imply that lineages produced by cells from the initial population grow and die independently of each other. This is a necessary simplification that is not biologically realistic, as cells usually compete for resources. A more biologically realistic model would include limited resources with competition between the lineages and spatial structure which plays an important role in the development of cancer \cite{martens2011spatial}. % add the description of the more realistic model here OK

@@

% TODO add subsection on the Gillespie simulations 
\subsection*{Gillespie simulations.} 
Simulations were performed using a Gillespie algorithm\cite{gillespie1976general,gillespie1977exact} programmed in Python. The wildtype population consisting of $N$ cells is assumed to be declining exponentially at the beginning of the simulations. The algorithm we utilized proceeds at time $t$ as follows:
\begin{enumerate}
\item First we draw the time step $\Delta t$ from an exponential distribution whose rate parameter is equal to the sum of the rates of all possible events.
\item Secondly we select a specific event $i$ according to its relative probability:
\begin{equation*}
p_i=\frac{\nu_i}{\sum \nu_j},
\end{equation*}
where $\nu_i$ is the rate of event $i$.
\item Finally, we update the population size and all the event rates for time $t+\Delta t$.
\end{enumerate}
We repeat steps 1 to 3 until either the population of cancer cells dies out or the mutant population reaches a level at which there is a $99\%$ probability that 
it will survive in perpetuity.
\paragraph{Data availability.} All code used to generate the figures in this article is available online at: {\color{blue}https://github.com/yoavram-lab/EvolutionaryRescue}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TODO combine all diagram figures to a single figure with multiple panels. and make it fit in a single page. OK

\begin{figure}
\begin{subfigure}{\textwidth}
\centering
\includegraphics[width=0.55\textwidth]{Figures/figureAneuploidy1.pdf}
\caption{}
\end{subfigure}

\bigskip

\begin{subfigure}{\textwidth}
\centering
\includegraphics[width=0.55\textwidth]{Figures/figureAneuploidy2.pdf}
\caption{}
\end{subfigure}

\bigskip

\begin{subfigure}{\textwidth}
\centering
\includegraphics[width=0.55\textwidth]{Figures/figureAneuploidy3.pdf}
\caption{}
\label{figureAneuploidy}
\end{subfigure}

\caption{$\left(\text{a}\right)$ Diagram of tumorigenesis by which a cell loses both of its tumor suppressing cells. The first tumor suppressing gene can be eliminated by a mutation at rate $2u$ or through chromosomal instability with rate $2p$. The last TSG can either be eliminated through mutation at rate $u$ or through chromosomal instability with rate $p$. Grey vertical lines represent active TSGs while the blue vertical lines represent inactivated TSGs. $\left(\text{b}\right)$ Diagram of the evolutionary rescue process which involves aneuploidy. The wildtype cell can either acquire trisomy at rate $u$ or acquire a fitness increasing mutation at rate $v$. Aneuploidy cells can  acquire a fitness increasing mutation at rate $v$. Finally, the aneuploidy cells can lose trisomy at rate $u$. $\left(\text{c}\right)$ Diagram of the evolutionary rescue model where the population of cancer cells is subdivided to wildtype euploid, wildtype aneuploid and mutant euploid cells that divide and die with rates $\lambda_w,  \lambda_a,  \lambda_m$ and $\mu_w, \mu_a, \mu_m$, respectively. Wildtype cells can acquire aneuploidy at rate $u$. Aneuploid and wildtype cells can mutate with rate $v$.}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\newpage
\section*{Results}

@@

We write \cref{survprobw} as
\begin{equation}\label{survprobwapprox1}
p_w=-\frac{1}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right).
\end{equation}
So, given an initial population of $N$ wildtype cells, the probability that the population will survive is given by
\begin{align}\label{aneuploidyresistentfirstapprox}
p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{v\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right] .
\end{align}
\Cref{SurvPlot} shows that the survival probability, $p_{est}$, quickly grows as the wildtype growth rate, $\lambda_w$, increases. % Todo I dont think this figure should show lambda_w>1, as in that case the wildtype is resistant and there is no rescue scenario. OK
\Cref{SurvPlotNData} show $p_{est}$ as a function of $N$, including comparison of our approximation \eqref{aneuploidyresistentfirstapprox} % ref to equation OK
and simulation results.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!t]
 \vspace*{1\baselineskip}
\includegraphics[width=1\textwidth]{Figures/P_est_growth.pdf}
\caption{Plot of the probability of survival of a population as a function of the proliferation rate of the wildtype cells. The blue line represents the exact solution \eqref{survprobw} and the orange line line represents the approximation \eqref{survprobwapprox1}. Here the population initially consists of $N$ wildtype cells and for the simulations we have chosen the following parameters: $N=10^4, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1.$}
\label{SurvPlot}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\paragraph{$\lambda_a<\mu_a$ and $\lambda_w<\mu_w$.}
We assume that $\lambda_a<\mu_a$ and $\lambda_w<\mu_w$ and, as a result, we rewrite \eqref{survprobw} and \eqref{survproba} as:
\begin{align*}
p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1-\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right),\\
p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right).
\end{align*}
\begin{align}\label{survprobwinitial}
p_w&\approx-\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
\nonumber
&\approx\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-v\right)}-\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ \label{survprobw2}
&\approx\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_w-\mu_w\right)}\left[\frac{u}{\left(\lambda_a-\mu_a\right)}-1\right],
\end{align}
where in the last line we have used the fact that $u,v\ll1$. Using the notational convention
\begin{equation}\label{notationalconv}
\Delta_i=\lambda_i-\mu_i,
\end{equation}
we write \eqref{survprobw} as
\begin{equation}\label{survprobwapprox2}
p_w=\frac{v\Delta_m}{\lambda_m\Delta_w}\left(\frac{u}{\Delta_a}-1\right).
\end{equation}
Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
\begin{align}\label{AneuploidyDeleteriousApprox}
p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{v\Delta_mN}{\lambda_m\Delta_w}\left(1-\frac{u}{\Delta_a}\right)\right],
\end{align}

@@

Given an initial population consisting of $N$ wildtype cancer cells, the probability that the population will survive is given by: 
\begin{align}\label{pestthirdcase}
p_{est}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w-u-v}\left(v\frac{\Delta_m}{\lambda_m}+\frac{u\left(\Delta_a-v\right)}{2\lambda_a}+u\sqrt{\frac{v\Delta_m}{\lambda_a\lambda_m}}\right)\right],
\end{align}
which we plot in Figure \ref{DeleteriousPlot} where we compare with numerical simulations and the exact result \eqref{survprobw}. The transition between the regimes defined by \eqref{AneuploidyDeleteriousApprox} and \eqref{pestthirdcase} respectively occurs at:
\begin{align}
\Delta_a^*=2vp_m+v+2\sqrt{vp_m\left(vp_m+\mu_a+v\right)}.
\end{align}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!t]
 \vspace*{1\baselineskip}
\includegraphics[width=1\textwidth]{Figures/SurvPlotNData.pdf}
\caption{Plot of the probability of survival of a population as a function of the initial population size of wildtype cells. The blue line represents the exact solution \eqref{survprobw}, the orange line line represents the approximation \eqref{aneuploidyresistentfirstapprox}, the green line represents the first order correction \eqref{survprobwapproxcorrected} and the red dots represents stochastic simulations. For the simulations we have chosen the following parameters: $\lambda_w=1-10^{-2}, \lambda_a=1+10^{-2},\lambda_m=1+10^{-1},\mu_w=1,\mu_a=1,\mu_m=1.$}
\label{SurvPlotNData}
\end{figure}
